---
editor_options:
  markdown:
    wrap: 72
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Including Plot

------------------------------------------------------------------------

# Prediction Assignment Write up

#### Mariana Martins

## Overview

This analysis meant to be the basis for the course quiz and a prediction
assignment write up. The main goal of the project is to predict the
manner in which 6 participants performed some exercise as described
below. This is the "class" variable in the training set. The machine
learning algorithm described here is applied to the 20 test cases
available in the test data and the predictions are submitted in
appropriate format to the Course Project Prediction Quiz for automated
grading.

## Data Loading and Exploratory Analysis

### Data Source

The training data for this project are available here:

#### [Training Set] <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:

#### [Test Set] <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

### Environment Setup

```{r message=FALSE, warning=FALSE}
library(knitr) 
library(caret) 
library(rpart) 
library(rpart.plot) 
library(rattle) 
library(randomForest) 
library(corrplot) 

set.seed(301)
```

### Data Loading and Cleaning

The next step is loading the dataset from the URL provided above. The
training dataset is then partitioned in 2 to create a Training set (70%
of the data) for the modeling process and a Test set (with the remaining
30%) for the validations. The testing dataset is not changed and will
only be used for the quiz results generation.

```{r}
TrainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv" 

TestUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" 

TrainFile<-"pml-traininig.csv" 

TestFile<-"pml-testing.csv"
```

Download the datasets

```{r}
if(!file.exists(TrainFile)) { 
  download.file(TrainUrl,destfile = TrainFile) 
  } 
training <- read.csv(TrainFile) 
if(!file.exists(TestFile)) {
  download.file(TestUrl,destfile = TestFile) 
  } 
testing <- read.csv(TestFile)
```

Create a partition using caret with the training dataset on 70,30 ratio

```{r}
inTrain <- createDataPartition(training$classe, p=0.7, list=FALSE)

TrainSet <- training[inTrain, ]

TestSet <- training[-inTrain, ]

dim(TrainSet) 
dim(TestSet)
```

Both created datasets have 160 variables. Let's clean NA, The Near Zero
variance (NZV) variables and the ID variables as well.

Remove variables with Nearly Zero Variance

```{r}
NZV <- nearZeroVar(TrainSet) 
TrainSet <- TrainSet[, -NZV] 
TestSet <- TestSet[, -NZV] 
dim(TestSet) 
dim(TrainSet)
```

Remove variables that are mostly NA

```{r}
AllNA <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95 
TrainSet <- TrainSet[, AllNA==FALSE] 
TestSet <- TestSet[, AllNA==FALSE] 
dim(TestSet) 
dim(TrainSet)
```

Remove identification only variables (columns 1 to 5)

```{r}
TrainSet <- TrainSet[, -(1:5)] 
TestSet <- TestSet[, -(1:5)] 
dim(TrainSet)
```

After cleaning, we can see that the number of variables for the analysis
are now only 53.

### Correlation Analysis

A correlation among variables is analysed before proceeding to the
modeling procedures.

```{r}
corMatrix <- cor(TrainSet[, -54]) 
corrplot(corMatrix, order = "FPC", method = "color", type = "lower", tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph
above. To make an even more compact analysis, a PCA (Principal
Components Analysis) could be performed as pre-processing step to the
datasets. Nevertheless, as the correlations are quite few, this step
will not be applied for this assignment.

## Prediction Model Building

Three popular methods will be applied to model the regressions (in the
Train dataset) and the best one (with higher accuracy when applied to
the Test dataset) will be used for the quiz predictions. The methods
are: Random Forests, Decision Tree and Generalized Boosted Model, as
described below. A Confusion Matrix is plotted at the end of each
analysis to better visualize the accuracy of the models.

### Random Forests

Model fit

```{r}
set.seed(301) 
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf", trContro =controlRF)
modFitRandForest$finalModel
```

Prediction on Test dataset

```{r}
predictRandForest <- predict(modFitRandForest, newdata=TestSet) 
confMatRandForest <- confusionMatrix(predictRandForest, as.factor(TestSet$classe)) 
confMatRandForest
```

Plot matrix results

```{r}
plot(confMatRandForest$table, col = confMatRandForest$byClass, main = paste("Random Forest - Accuracy =", round(confMatRandForest$overall['Accuracy'], 4)))
```

### Decision Tree

Model fit

```{r}
set.seed(301) 
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class") 
fancyRpartPlot(modFitDecTree)
```

Prediction on Test dataset

```{r}
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class") 
confMatDecTree <- confusionMatrix(predictDecTree, as.factor(TestSet$classe) )
confMatDecTree
```

Plot matrix results

```{r}
plot(confMatDecTree$table, col = confMatDecTree$byClass, main = paste("Decision Tree - Accuracy =", round(confMatDecTree$overall['Accuracy'], 4)))
```

### Generalized Boosted Model

Model fit

```{r}
set.seed(301) 
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1) 
modFitGBM <- train(classe ~ ., data=TrainSet, method = "gbm", trControl = controlGBM, verbose = FALSE)

modFitGBM$finalModel
```

Prediction on Test dataset

```{r}
predictGBM <- predict(modFitGBM, newdata=TestSet) 
confMatGBM <- confusionMatrix(predictGBM, as.factor(TestSet$classe) )
confMatGBM
```

Plot matrix results

```{r}
plot(confMatGBM$table, col = confMatGBM$byClass, main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

## Applying the selected model

The accuracy of the 3 regression modeling methods above are:

-   Random Forest : 0.9986

-   Decision Tree : 0.7295

-   GBM : 0.989

In that case, the Random Forest model will be applied to predict the 20
quiz results (testing dataset) as shown below.

```{r}
predictTEST <- predict(modFitRandForest, newdata=testing) 
predictTEST
```
